Dynamic Stochastic Approximation Monte Carlo

### 2018-07-12 papers/histogram/sad.tex
edited results to reflect our analysis of translation scale and minT.

For the square-well fluid, our simulations treat systems with a
well-width of $\text{ww} = 1.30$ and a filling fraction of either
$\text{ff} = 0.10, 0.20,$ or $0.30$.  The simulation explores the
energy space of the system down to a reduced temperature of
$T_{\text{min}} = 0.2$.  All simulations lead to the minimum important
energy $E_{\text{min}}$ and maximum entropy state $S_{\text{max}}$
being calculated (with the exception of the WL-TMMC method where both
of these parameters are needed a priori).  In the sections below, we
examine systems with different particles numbers and filling fractions.

We use the average entropy error vs moves a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a `golden' standard. For each simulation, the `golden' standard is
SAMC for the best choice of $t_0$ for $10^{11}$ moves followed by TMMC.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{golden}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.1-avg-error}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_10-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.10 with
  SAMC followed by TMMC serving as the `golden' standard.}\label{fig:N50-ff0.1-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_20-N50-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.20 with TMI acting as the
  `golden' standard.}\label{fig:N50-ff0.2-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.30 with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.30 with TMI acting as the
  `golden' standard.}\label{fig:n500-ff0.2}
\end{figure}

\subsubsection{A periodic system with 10\% filling fraction and $(N = 50)$}
In order to determine how the various methods scale with size,
simulations are run for systems with different particle numbers.
Figure~\ref{fig:N50-ff0.1-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a filling fraction of 0.10.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

\subsubsection{A periodic system with 20\% filling fraction and $N = 50$}

Figure~\ref{fig:N50-ff0.2-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a filling fraction of 0.20.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

\subsubsection{A periodic system with 30\% filling fraction and $N = 50$}

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a filling fraction of 0.30.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

\subsubsection{A periodic system with 30\% filling fraction and $N = 500$}

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\text{ff} = 0.30$ is chosen which yields $N = 500$
particles in a cubical system.  These particles are randomly inserted
into the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05$).

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed that all the methods from
this family struggle to converge if the update factor decreases faster
than $\nicefrac{10^5}{\sqrt{t}}$. Fig.~\ref{fig:n500-ff0.2} shows both
WL and SAMC with inappropriate choices of $t_0$ i.e. less than
$\nicefrac{10^5}{\sqrt{t}}$ failing to converge in a reasonable amount
of time. Thus, the update factor $\gamma$, shown as a function of
simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves useful in
determining which methods will converge in a reasonable amount of time.
We compare each method to a `golden' standard which we define to be an
ideal choice of SAMC for $10^{11}$ moves followed by TMMC.  This gives an
unbiased comparison with our developed method SAD. The dynamic version
of SAMC which we call SAD performs as well as SAMC (assuming the user
can correctly identify an appropriate choice of $t_0$).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

>>>>

For the square-well fluid, our simulations treat systems with a
well-width of $\lambda = 1.30$ and a filling fraction of
$\eta = 0.30$. We also explore various translation scales $\delta_0 = 0.005,
0.05,$ and $0.5$. The simulation explores the
energy space of the system at a reduced temperature of
$T_{\text{min}} = 0.2$ and $0.5$.  All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine systems parameters on the convergence of the methods.

We use the average entropy error vs moves a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a `golden' standard. For each simulation, the `golden' standard is
SAMC for the best choice of $t_0$ for $10^{11}$ moves followed by TMMC.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{golden}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.3-avg-error}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for $\lambda = 0.30$ with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error.pdf}
  \caption{The average entropy error is shown for $\lambda = 0.30$ with TMI acting as the
  `golden' standard.}\label{fig:n500-ff0.3}
\end{figure}

~\jpsays{We have discussed including a figure for the $\lambda = 30\%$ and $N = 50$
showing translation scale impact. Also a figure for the $\lambda = 30\%$ and $N = 50$
showing the impact of choice of minimum temperature.}

\subsection{A periodic system with $\lambda = 30\%$ and $N = 50$}

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a $\lambda = 0.30$.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

\subsubsection{translation scale}
For the $\lambda = 30\%$ and $N = 50$ system, we explore the choice of $\delta_0$
as a parameter. Smaller translation scales should cause the given method to take
additional time to explore all energies.  A larger translation scale could result 
in the simulation taking shorter or longer (because longer distance moves may end up
not being allowed). Ideal methods should scale linearly with the $\delta_0$, that is,
a decrease in the translation scale should result in a proportional increase in convergence
time. Some methods such as WL may not be able to handle arbitrary $\delta_0$.

\subsubsection{minimum temperature}
For the $\lambda = 30\%$ and $N = 50$ system, we explore the choice of $T_{\min}$
as a parameter. We have theoretically postulated that a dynamic version of SAMC which
we call SAD should depend upon this physically meaningful parameter. We thus explore
the behaviour of SAD for two different reduced temperatures.

\subsection{A periodic system with $\lambda = 30\%$ and $N = 500$}

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\lambda = 0.30$ is chosen which yields $N = 500$
particles in a cubical lattice system.  These particles are randomly inserted
into the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05$).

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed that all the methods from
this family struggle to converge if the update factor decreases faster
than $\nicefrac{10^5}{\sqrt{t}}$. Fig.~\ref{fig:n500-ff0.3} shows both
WL and SAMC with inappropriate choices of $t_0$ i.e. less than
$\nicefrac{10^5}{\sqrt{t}}$ failing to converge in a reasonable amount
of time. Thus, the update factor $\gamma$, shown as a function of
simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves useful in
determining which methods will converge in a reasonable amount of time.
We compare each method to a `golden' standard which we define to be an
ideal choice of SAMC for $10^{11}$ moves followed by TMMC.  This gives an
unbiased comparison with our developed method SAD. The dynamic version
of SAMC which we call SAD performs as well as SAMC (assuming the user
can correctly identify an appropriate choice of $t_0$).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.
